---
title: "Frequency of Type 2 Diabetes Among Women of Pima Indigenous Heritage Living Near Phoenix, Arizona"
output: html_document
date: "2023-07-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing Libraries

This chunk installs the required libraries to form the chats needed to explain the data for the final project.

```{r packages}
if(!require(pacman))install.packages("pacman")

pacman::p_load('dplyr', 'tidyr', 'gapminder',
               'ggplot2',  'ggalt',
               'forcats', 'R.utils', 'png', 
               'grid', 'ggpubr', 'scales',
               'bbplot')

library(ggplot2)
library(dplyr)
```

## Diabetes DataFrame

The diabetes dataframe comes from Datahub.io. It appears in Kaggle and many research papers regarding the frequency of type 2 diabetes among indigenous Native American people. The columns included in the table are:
Pregnancies - Number of pregnancies.
GlucoseLevel - Plasma glucose concentration from a 2-hour oral glucose tolerance test.
BloodPressure - Diastolic blood pressure measurement (mm Hg) taken at the time of the study.
SkinThickness - Triceps skin fold thickness (mm).
Insulin - 2-Hour serum insulin (mu U/ml).
BMI - Body Mass Index (weight in kg/(height in m)^2).
DiabetesPedigreeFunction - A measure of genetic influence that provides data on diabetes mellitus history in relatives and the genetic relationship of those relatives to the patient.
Age - The woman's age in years.
Outcome - Tested either positive or negative for diabetes.

```{r read file, echo=FALSE}
file = "https://datahub.io/machine-learning/diabetes/r/diabetes.csv"
diabetesData <- read.csv(file)

# Replaces 0s in all columns with NAs except for preg and class.
diabetesData$plas[diabetesData$plas == 0] <- NA
diabetesData$pres[diabetesData$pres == 0] <- NA
diabetesData$skin[diabetesData$skin == 0] <- NA
diabetesData$insu[diabetesData$insu == 0] <- NA
diabetesData$mass[diabetesData$mass == 0] <- NA
diabetesData$pedi[diabetesData$pedi == 0] <- NA

# Drops rows with NAs in all columns except for preg and class.
diabetesData <- na.omit(diabetesData)

# Renames columns for easier readability
diabetesData <- data.frame(Pregnancies=diabetesData$preg, GlucoseLevel=diabetesData$plas, BloodPressure=diabetesData$pres, SkinThickness=diabetesData$skin, Insulin=diabetesData$insu, BMI=diabetesData$mass, DiabetesPedigreeFunction=diabetesData$pedi, Age=diabetesData$age, Outcome=diabetesData$class)

head(diabetesData)
```

## Frequency of Being Diagnosed with Diabetes Based on Age

Plot 1 measures the frequency of type 2 diabetes in women of Pima indigenous heritage living near Phoenix, Arizona as they age. Older women are more likely to be tested positive for type 2 diabetes than younger women.

```{r age diabetic plot, echo=FALSE}
ggplot(diabetesData, aes(x=Outcome, y=Age)) + geom_point(position="Jitter", width=0.25, alpha=0.5) + labs(y="Age", x="Diabetes Status", title="Frequency of Diabetes by Age", subtitle="Pima indigenous women near Phoenix, AZ") + theme(legend.position = "top", legend.justification = "left") + bbc_style() + theme(axis.title = element_text())
```
## Frequency of Diabetes Between Age and Number of Pregnancies

Plot 2 measures the frequency of type 2 diabetes in women of Pima indigenous heritage living near Phoenix, Arizona as they age or have more children. Very few women have 5 or more children because nowadays, families do not produce 5 or more children. Women get pregnant at a younger age because these years are their productive years. Most young women, whether they are pregnant or not, are less likely to become diabetic. As the woman ages, her chance of becoming diabetic increases. Women can become diabetic regardless of the number of pregnancies.

```{r age pregnancies, echo=FALSE}
ggplot(diabetesData, aes(x=Pregnancies, y=Age, color=Outcome)) + geom_point(position="Jitter", width=0.25, alpha=0.5) + labs(y="Age", x="Number of Pregnancies", title="Diabetes by Age & Pregnancy", subtitle="Pima indigenous women near Phoenix, AZ") + theme(legend.position = "top", legend.justification = "left") + bbc_style() + theme(axis.title = element_text()) + scale_colour_manual(values = c("#2D9900", "red"))
```

## Frequency of Diabetes Between Age and Blood Pressure Level

Plot 3 measures the frequency of type 2 diabetes between age and blood pressure. Younger women under 30 years old usually test negative for type 2 diabetes. Their diastolic blood pressure is within the normal limit of 60 to 80 mm Hg. Around 35 years old to 60 years old, the chances of becoming diabetic and hypertensive increases.

```{r age blood pressure plot, echo=FALSE}
ggplot(diabetesData, aes(x=BloodPressure, y=Age, color=Outcome)) + geom_point(position="Jitter", width=0.25, alpha=0.5) + labs(y="Age", x="Diastolic Blood Pressure Level (mm Hg)", title="Diabetes Status by Age & BP", subtitle="Pima indigenous women near Phoenix, AZ") + theme(legend.position = "top", legend.justification = "left") + bbc_style() + theme(axis.title = element_text()) + scale_colour_manual(values = c("#2D9900", "red"))
```
## Frequency of Diabetes Between BMI and Glucose Level

Plot 4 measures the frequency of type 2 diabetes between glucose level and BMI. As women's BMI reaches 25 and up (Obesity), their glucose level increases. The higher the BMI above the normal range (18.5–24.99), the higher the chance the woman becomes diabetic. According to American Diabetes Association (https://diabetes.org/diabetes/a1c/diagnosis), the normal glucose level in an oral glucose tolerance test is less than 140 mg/dL. Prediabetes is 140 to 199 mg/dL. Diabetes is 200 mg/dL or higher. At about 130 mg/dL glucose level and higher, majority of women are already at a high risk of becoming diabetic.

```{r bmi glucose level plot, echo=FALSE}
ggplot(diabetesData, aes(x=GlucoseLevel, y=BMI, color=Outcome)) + geom_boxplot(outlier.shape=NA) + geom_jitter(alpha=0.5) + labs(y="BMI", x="Glucose Level in an Oral Glucose Tolerance Test (mg/dL)", title="Diabetes by BMI & Glucose Level", subtitle="Pima indigenous women near Phoenix, AZ") + theme(legend.position = "top", legend.justification = "left") + bbc_style() + theme(axis.title = element_text()) + scale_colour_manual(values = c("#2D9900", "red"))
```


## Frequency of Diabetes in Different Tricep Skin Fold Thickness and BMI

Plot 5 measures the frequency of type 2 diabetes between skin thickness and BMI. As a woman's BMI increases and her triceps skin fold thickness increases, the likelihood of becoming diabetic is prominent. Women with BMI 25 and higher and have thicker triceps skin folds (mm) are at a higher risk of getting type 2 diabetes.

```{r bmi skin thickness plot, echo=FALSE}
ggplot(diabetesData, aes(x=BMI, y=SkinThickness, color=Outcome)) + geom_boxplot(outlier.shape=NA) + geom_jitter(alpha=0.5) + labs(y="Triceps Skin Fold Thickness (mm)", x="BMI", title="Diabetes by BMI & Skin Thickness", subtitle="Pima indigenous women near Phoenix, AZ") + theme(legend.position = "top", legend.justification = "left") + bbc_style() + theme(axis.title = element_text()) + scale_colour_manual(values = c("#2D9900", "red"))
```
## Frequency of Diabetes in Family History

Plot 6 measures the frequency of type 2 diabetes in the diabetes pedigree function. The diabetes pedigree function is a function that measures the likelihood of type 2 diabetes based on family history. A higher value in the diabetes pedigree function indicates that the woman is at risk of developing type 2 diabetes. Women can still become diabetic even when their diabetes pedigree is lower than 0.5. However, there are more women who tested negative for diabetes with a diabetes pedigree of 0.5 and lower.

```{r diabetes pedigree function, echo=FALSE}
ggplot(diabetesData, aes(x=Outcome, y=DiabetesPedigreeFunction)) + geom_boxplot(outlier.shape=NA) + geom_jitter(alpha=0.5) + labs(y="Diabetes Pedigree Family History", x="Diabetes Status", title="Diabetes Status by Family History", subtitle="Pima indigenous women near Phoenix, AZ") + theme(legend.position = "top", legend.justification = "left") + bbc_style() + theme(axis.title = element_text()) + scale_colour_manual(values = c("#2D9900", "red"))
```

## Association Rules For Diabetes Frequency

The Association Rules from the apriori library compare the risk of type 2 diabetes occurring under different body conditions, such as number of Pregnancies, Blood Glucose Level, Blood Pressure, Skin Thickness, Insulin level, BMI, and Age.

```{r apriori, echo=FALSE}
set.seed(3)
appearance = list(default = "lhs", rhs = ("Outcome=tested_positive"))
parameter = list(supp = 0.001, conf = 0.8)

library(arules)
library(arulesViz)

# Converts data frame to transactions.
trans <- as(diabetesData, "transactions")

#Uses apriori() function on diabetic people.
rules <- apriori(trans, parameter = parameter, appearance = appearance)

# Views rules.
inspect(rules)
```
The arulesViz library visualizes the results. The x axis represents the support while the y axis represents the confidence or lift. Support indicates how frequently a set of items appear in the data. It is the number of transactions that contains this set. Confidence or lift indicates how often the support rule is true. Lift is a measure of association using both support and confidence. The lift value of an association rule includes the list of rules whose confidence is above the threshold.

```{r}
# BEGIN SOLUTION
library(arulesViz)

plot(rules, main = "Visualized Association Rules Diabetes in Pima Indigenous Women")
```
The previously created Association Rules can be sorted and filtered to find the best rules. A high filter, such as lift > 2.9, picks out the best rules. The best rules have a very high confidence (precision) and low support (recall) value because they show that the data is very precise. In this data, values with a lift > 2.9 show that certain table columns have a strong association with positive type 2 diabetes diagnosis.
```{r}
# BEGIN SOLUTION
# Sorts the list of previously created rules.
sortedRules <- sort(rules)

# Uses lift > 2.9 to find the best rules.
subrules <- subset(sortedRules, lift>2.9)
plot(subrules, main = "Best Association Rules Diabetes in Pima Indigenous Women")
```